Rice yield improvements through plant breeding are offset by inherent yield declines over time

### 2018-03-11 output/review_responses.tex

\textbf{Response:} The term ``yield erosion'' has been removed due to
these concerns. However, we have taken great pains to specifically
point out that the evidence suggests this inherent to the system (as
the title of this manuscript clearly states), that it is similar to
observations from other rice systems (lines 211-217), and that it is
plausibly due to the environmental aspects the reviewer lists (lines
190-210).

However, the evidence from this study explicitly states that given
these data and this analysis, a rate of improvement of 29 kg/ha/yr is
unlikely. The reviewer appears to have missed one of the main
conclusions of the manuscript. We have edited for greater clarity. \\

\textbf{Comment: Really it all depends on the definition of potential
  yield. See van Ittersum and colleagues in FCR Vol 143 March
  2013. Its determined by latest best yielding Cv and the current E in
  the absences of any manageable stresses, both biotic and abiotic,
  and measured in good experiments and models based on such
  experiments. A breakdown in cv resistance is probably best described
  as “erosion”, while a deterioration in the E due, for example, to a
  negative climate trend, is a decrease in the potential yield, but a
  new cv tolerating this deterioration would lift the potential yield
  (see also below).}\\

\textbf{Response:} This is a good point and we have in part addressed
this issue by removing the term “yield erosion”. We have also,
addressed the issue raised by van Ittersum that a changing climate may
change yield potential (lines ).\\

@@

\textbf{Comment: The authors should read McKay et al (2010, TAG 122,
  225-238) where a similar situation in winter wheat in the UK is
  described. But by having plots with and without fungicide over the
  last 40 years or so, these authors could measure the erosion and
  narrow down the cause (foliar disease), which did vary with
  cultivar.}\\

\textbf{Response:} This is an interesting paper (assuming the reviewer
intended \textbf{Mackay et al. 2011}), but unfortunately these data do
not lend themselves to similar explorations. Unlike Mackay et al., we
do not have paired treated/untreated plots. In the course of writing
this paper, we discussed various means to to narrow down the cause of
yield losses, but ultimately lack the data to do so. We have added a
citation to McKay et al. to the manuscript.\\

\textbf{Comment: Also the authors need to refer to McKenzie et al
  (2014) in the CSSA Special Publicn 33 with a whole section on rice
  yield progress in Ca.}\\

\textbf{Response:} We have addressed this paper more
thoroughly (lines ). Both McKenzie et al. (2014) and Samonte et
al (2014) fail to take yield declines into account in their analyses
and conclusions.  When we exclude these effects, we estimate similar
effects. We show that accounting yield declines results in better
explanation of the data, and resolves appearant contradiction between
short-term and long-term data. \\

### 2018-03-11 output/breed_paper_revised.Rnw
Modifies end of paper.
Updates with minor edits.

Meeting the challenge of feeding a growing population with limited
resources will require increasing the yield potential of staple crops,
such as rice. Yet many high-yielding, intensive production systems
have experienced slow rates of yield improvement in recent years
despite a demonstrated increase in the yield potential of new crop
cultivars. We analyzed experimental data from one such cropping
system, i.e., California (CA) rice, in order to quantify improvements
made in the genetic yield potential obtained through plant
breeding. California rice systems are among the highest in the world
and close to maximum yield potential. Specifically, the hypothesis was
tested that if rice cultivar yields decline over time then apparent
yield increases in side-by-side yield comparison tests will not
reflect increases in yield potential. This hypothesis was tested using
33 years of experimental yield data from the California Cooperative
Rice Research Foundation Rice Experiment Station. Based on
side-by-side comparisons of old and new rice cultivars which do not
consider yield decline over time, there was an apparent increase in
yield. However, the yields of older cultivars declined at an estimated
rate of 29.3 kg ha\tsups{-1} year\tsups{-1} (90\% credible interval
-4.4 to -53.3) after initial selection. Once this effect was
considered, the yield advantage of newer cultivars over old was
uncertain (-3.3 kg ha\tsups{-1} year\tsups{-1}, 90\% credible interval
-36.1 to 31.5). These results highlight 1) the importance of
continuous crop improvement and deployment of new cultivars simply to
maintain existing yields, and 2) to increase the genetic yield
potential, higher yield targets are needed. Importantly, when breeding
near the yield potential, despite the limited yield gains, significant
advances in improving quality and reducing crop duration have been
made.

@@

Certainly one question in looking at this situation is, ``do new
cultivars have greater yield potential''?  Like many breeding
programs, one of the goals of the CCRRF is to increase yields through
breeding cultivars with increased yield potential. However, when the
popular cultivar M-202 was released in 1985, it was reported to yield
an average of 11 Mg ha\tsups{-1}~\citep[14\% moisture content
(MC);][]{m202}. In 2015, the reported average yield of the newest
released cultivar M-209 was 10.8 Mg ha\tsups{-1}~\citep[14\%
MC;][]{m209}. However, in annual breeding reports and cultivar release
announcements, data based on yield performance in side-by-side cultivar
trials show that new cultivars usually out yield the older
cultivars, typically by 3--5\% (e.g., M-209 release description at \url{http://www.crrf.org/linked/2015annualreport.pdf}). In other rice systems, a decrease in a cultivar's yield
performance over time has been observed~\citep{Peng1998, Peng2010,
Peng2000, Datta1995}, sometimes termed ``yield erosion''~\citep{Peng1998}. It
has been speculated that this yield decline over time is due to the inability of
a cultivar to adapt to changing biotic and abiotic
conditions~\citep{Peng1998, Peng2010, Peng2000, Datta1995}. Yield
declines over time may explain the apparent contradiction above, in which
side-by-side yield comparisons show yield improvements in cultivars
while overall there may be little change in yield potential. To
address this issue we tested the hypothesis that if rice cultivar
yields erode over time then apparent yield increases in side-by-side
yield comparison tests will not reflect the true magnitude of yield
changes over time.

@@

\begin{equation}
\begin{split}
 yield_{ijk} = Intercept + (Rate_{yield\ decline}* {years\ since\
   release\ ik}) \\+ (Rate_{yield\ improvement}* {release\ year\ ik})\\ 
 + \beta_{year\ ij} + \beta_{cultivar\ ik} +
 \beta_{cultivar\ ik:year\ ij} + \epsilon \\
\end{split}
\end{equation}

Where $yield_i$ is the grain yield for cultivar $k$ in plot $i$ for
year $j$,
the Intercept is the average yield for the RES across cultivars and years, $Rate_{yield\
decline}$ is rate of yield decline per ${years\ since\ release\ i}$
(mean-centered, i.e., negative for years prior to release, zero at the year of release),
$Rate_{yield\ improvement}$ is the rate of yield improvement per
year (${release\ year\ i}$), $\beta_{year}$ is the random effect of
year, and $\beta_{cultivar}$ is the random effect of cultivar. This
full model was modified to further explore yield dynamics over
time. For comparison to estimates where yield decline over time is not taken
into account, the model was used as described above except omitting
the $Rate_{yield\ decline\ i}$ term (Model 2). Lastly, a model was fit
to estimate if there were differences between cultivars in the rate of
yield decline over time by adding a random slope by cultivar (Model 3).

@@

\begin{figure}[H]
<<echo = FALSE, fig = TRUE, width = 7, height = 6>>=
library(ggplot2)
library(ggthemes)
ggplot(mod_data, aes(x = year, y = yield_kg/1000)) +
    geom_point(size = 0.5, alpha = 0.5) +
    geom_smooth(method = "lm", color = "black") +
    theme_few() +
    xlab("Year") +
    ylab(expression("Grain yield ( Mg ha"^-1~")")) + 
    facet_wrap(~id_fact, ncol = 4)

@@

In CA, the cultivars with broad adaptation and adoption over the
33-year period of this study are M-202, M-206, and M-105. Cold
tolerance is a major problem for much of the California rice producing
region ~\citep{espe2017}, although it is a bigger problem in the
southern portion of the Sacramento Valley due to the influence of the
ocean winds through the Delta ~\citep{espe16a}. These three cultivars
have a higher degree of cold tolerance, high yield potential, and
relatively short duration and are thus broadly grown with over 70\% of
the area in these cultivars at any given time during the study
period~\citep{ucce}. There are several cultivars that were found to
have better than average yields, even after correcting for yield
declines over time. However, these cultivars (M-201, M-204, M-205, and possibly
M-209; Fig. \ref{fig:ids_comp})) are not widely adopted. Importantly,
not all medium grain cultivars that were released were purported to
have higher yields than previous cultivars. Some cultivars had special
traits, but upon release had similar yields to more broadly adapted
cultivars. Examples of this include M-208, which had resistance to
blast disease (\textit{Magnaporthe grisea}) and M-402 which is a
premium quality cultivar. Other cultivars such as M-205 and M-209,
while high yielding, do not have the cold tolerance of the other
cultivars and are thus only recommended for the northern portion of
the Sacramento Valley.

@@

The high yielding variety M-202 was released in 1985 and became the
most widely planted cultivar for two decades~\citep{mkenzie2014}. At
the time of release, M-202 yields were reported at 11.0 Mg
ha\tsups{-1}~\citep{m202} which is almost identical to what we . This
value of 11.0 Mg ha\tsups{-1} is what we also found in this analysis
based on the average experimental yields prior to public release
(Fig. \ref{fig:first_yrs}). Since the release of M-202, the rice
yields of newly released varieties cultivars have remained between
10.8 and 11.5 Mg ha\tsups{-1} (Fig. \ref{first_yrs}), with the
exception of M-204 and M-205 (M-204 is no longer grown and M-205 has
limited adoaptation range so is only grown in a limited area). The
variety cultivar M-206 was released in 2004 and is currently still the
most widely grown rice varietycultivar in California. At the time of
its release, M-206 was reported to yield \Sexpr{round(9350*1.12,1)} Mg
ha\tsups{-1} -- also similar to what we found based on our analysis
(Fig. \ref{fig:first_yrs}). Based on this premise, it is clear that
improvements in yield potential have not been made over the 33-yr
period of this study. In contrast, both~\citet{Omar2014} and
~\citet{mckenzie2014}, report yield potential gains in California rice
systems based on analyses that showed that newly released cultivars
had higher yields than previously released cultivars. Based on their
analyses,~\citet{Omar2014} reported yield gains from the rice breeding
program of 37 to 60 kg ha\tsups{-1} yr\tsups{-1} and
~\citet{mckenzie2014} a yield increase of 25 kg ha\tsups{-1}
yr\tsups{-1}. Using a similar approach to~\citet{Omar2014} and
~\citet{mckenzie2014}, when the compared average yields for the period
2011--2016 of varieties cultivars released since 1982 (and still in
the variety trials from 2011-- 2016) was compared without accounting
for yield declines over time, yields were estimated to increase by
\Sexpr{only_release_coef} kg ha\tsups{-1} yr\tsups{-1}
(Fig. \ref{fig:wo_erosion}), almost identical to what
~\citet{mckenzie2014}.

The only way to reconcile the two apparently contradictory findings,
is that the yield of potential of varieties cultivars decline over
time such that when newly released varieties cultivars are compared
with the older variety cultivar (as shown by ~\citet{Omar2014},
~\citet{mckenzie2014}, and Fig. \ref{fig:wo_erosion}), yields of the
new cultivar variety are higher. Here we provide evidence that
cultivar varietal yield does decline - on average at a rate of
\Sexpr{yrs_coef} kg ha\tsups{-1} yr\tsups{-1}
(Fig. \ref{w_erosion}). ~\citet{Peng2010} also found that a cultivar’s
varieties yield potential declines over time and reported a 15\%
decline over a 30 yr period.  In this study, over a 30 year period and
a \Sexpr{yrs_coef} kg ha\tsups{-1} yr\tsups{-1} yield decline rate,
the yield decline would be approximately
\Sexpr{round((yrs_coef*-30)/mean(mod_data$yield_kg), 3)*100}\%.

This analysis suggests that the CCRRF breeding program has
maintained but not substantially improved the yield potential in the
face of changing biotic and abiotic pressures over the 33 year time
frame of this study. This conclusion holds
true even when evaluating the cultivars which have been breed for
broad adaption and high yield potential (M-202, M-206 and
M-105).

@@

Yields of a particular cultivar may decline overtime for a number of
reasons. First, ~\citet{Nie2009} reported yield declines in rice due
to continuously flooded conditions over a long time period which could
be overcome with improved N management. This is likely not the case
here as the fields used in this study grow rice only one season a year
and tend to be non-flooded for the remainder of the year. Furthermore,
at this location and unlike most CA rice fields, the fields are left
fallow every other year. Secondly, small changes in biotic and abiotic
forces have previously been speculated as the cause for yield decline
over time of the cultivar IR8 at IRRI~\citep{Peng1998, Peng2000,
  Peng2010}. Since the cultivar's genetics remain stable from the time
of selection but the conditions the cultivar experiences over time are
continuously changing, accumulated changes too small to reliably
measure could plausibly lead to yield declines over
time~\citep{Peng1998, Peng2000, Peng2010, Mackay2011}. However, it is
not clear if this mechanism is responsible for the yield decline over
time of \Sexpr{yrs_coef} Mg ha\tsups{-1} year\tsups{-1} found in this
analysis. Biotic stressors such as pests and diseases are low in this
Mediterranean climate. Furthermore, these fields are fallowed every
other year and the residues are burned, further lowering the potential
for pest and disease pressure. Thirdly, changing climate may influence
yield potential with either warmer~\citep{van2013} or
cooler~\citep{espe2017} temperatures reducing
yields. Yield potential can be defined as the yield of optimum
cultivar in the absence of biotic and abiotic stresses in a given
environment \citep{vanIttersum2013}. However, ~\citet{espe2017} found
no substantial trends in growing season temperatures (either high,
low, or average temps) in the Sacramento Valley during the study
period, suggesting that older cultivars do not have lower yield
potential due to changing climatic factors. While our analysis
indicates that there is no interaction between cultivar and year (CI
includes zero), visual examination suggests that cultivars may be
different in terms of rates of yield decline
(Fig. \ref{fig:all_cult}); however, we do not have sufficient data nor
the proper controls in the existing data set to fully test this
hypothesis. More data is needed to further explore the exact cause of
the yield declines observed here, as well to determine if different
cultivars have differential declines over time.

>>>>

Meeting the challenge of feeding a growing population with limited
resources will require increasing the yield potential of staple crops,
such as rice. Yet many high-yielding, intensive production systems
have experienced slow rates of yield improvement in recent years
despite a demonstrated increase in the yield potential of new crop
cultivars. We analyzed experimental data from one such cropping
system, i.e., California (CA) rice, in order to quantify improvements
made in the genetic yield potential obtained through plant
breeding. California rice systems are among the highest in the world
and close to maximum yield potential. Specifically, the hypothesis was
tested that if rice cultivar yields decline over time then apparent
yield increases in side-by-side yield comparison tests will not
reflect increases in yield potential. This hypothesis was tested using
33 years of experimental yield data from the California Cooperative
Rice Research Foundation Rice Experiment Station. Based on
side-by-side comparisons of old and new rice cultivars which do not
consider yield decline over time, there was an apparent increase in
yield. However, the yields of
older cultivars were found to decline at an estimated rate of 29.3 kg ha\tsups{-1}
year\tsups{-1} (90\% credible interval -4.4 to -53.3) after initial
selection. Once this effect was considered, the yield advantage of
newer cultivars over old was uncertain (-3.3 kg ha\tsups{-1}
year\tsups{-1}, 90\% credible interval -36.1 to 31.5). These results
highlight 1) the importance of continuous crop improvement and
deployment of new cultivars simply to maintain existing yields, and 2)
to increase the genetic yield potential, higher yield targets are
needed. Importantly, when breeding near the yield potential, despite
the limited yield gains, significant advances in improving quality and
reducing crop duration have been made.

@@

Certainly one question in looking at this situation is, ``do new
cultivars have greater yield potential''?  Like many breeding
programs, one of the goals of the CCRRF is to increase yields through
breeding cultivars with increased yield potential. However, when the
popular cultivar M-202 was released in 1985, it was reported to yield
an average of 11 Mg ha\tsups{-1}~\citep[14\% moisture content
(MC);][]{m202}. In 2015, the reported average yield of the newest
released cultivar M-209 was 10.8 Mg ha\tsups{-1}~\citep[14\%
MC;][]{m209}. However, in annual breeding reports and cultivar release
announcements, data based on yield performance in side-by-side cultivar
trials show that new cultivars usually out yield the older
cultivars, typically by 3--5\% (e.g., M-209 release description at \url{http://www.crrf.org/linked/2015annualreport.pdf}). In other rice systems, a decrease in a cultivar's yield
performance over time has been observed~\citep{Peng1998, Peng2010,
Peng2000, Datta1995}. It
has been speculated that this yield decline over time is due to the inability of
a cultivar to adapt to changing biotic and abiotic
conditions~\citep{Peng1998, Peng2010, Peng2000, Datta1995}. Yield
declines over time may explain the apparent contradiction above, in which
side-by-side yield comparisons show yield improvements in cultivars
while overall there may be little change in yield potential. To
address this issue we tested the hypothesis that if rice cultivar
yields erode over time then apparent yield increases in side-by-side
yield comparison tests will not reflect the true magnitude of yield
changes over time.

@@

\begin{equation}
\begin{split}
  yield_{ijk} = Intercept + (Rate_{yield\ decline}* {years\ since\ release_{ik}}) \\
  + (Rate_{yield\ improvement}* {release\ year_{k}}) \\
  + \beta_{year_j} + \beta_{cultivar_k} + \beta_{cultivar_k:year_j} + \epsilon \\
\end{split}
\end{equation}

Where $yield_i$ is the grain yield for cultivar $k$ in plot $i$ during
year $j$,
the Intercept is the average yield for the RES across cultivars and years, $Rate_{yield\
decline}$ is the rate of yield decline per
${years\ since\ release_{ki}}$ for cultivar $k$ in plot $i$
(mean-centered, i.e., negative for years prior to release, zero at the year of release),
$Rate_{yield\ improvement}$ is the rate of yield improvement per
year (${release\ year_{ik}}$), $\beta_{year_j}$ is the random effect of
year $j$, and $\beta_{cultivar_k}$ is the random effect of cultivar
$k$, and lastly $\beta_{cultivar_k:year_j}$ is the interaction term
for cultivar $k$ in year $j$. This
full model was modified to further explore yield dynamics over
time. For comparison to estimates where yield decline over time is not taken
into account, the model was used as described above except omitting
the $Rate_{yield\ decline}$ term (Model 2). Lastly, a model was fit
to estimate if there were differences between cultivars in the rate of
yield decline over time by adding allowing the rate of decline to vary
by cultivar (i.e., adding
a $Rate_{yield\ decline_k} * years\ since\ release_{ki}$ term to the
model for the rate of yield decline for cultivar $k$) (Model 3).

@@

\begin{figure}[H]
<<echo = FALSE, fig = TRUE, width = 7, height = 6>>=
library(ggplot2)
library(ggthemes)
ggplot(mod_data, aes(x = year, y = yield_kg/1000)) +
    geom_point(size = 0.5) +
    geom_smooth(method = "lm", color = "black") +
    theme_few() +
    xlab("Year") +
    ylab(expression("Grain yield ( Mg ha"^-1~")")) + 
    facet_wrap(~id_fact, ncol = 4)

@@

In CA, the cultivars with broad adaptation and adoption over the
33-year period of this study are M-202, M-206, and M-105. Cold
tolerance is a major problem for much of the California rice producing
region ~\citep{espe2017}, although it is a bigger problem in the
southern portion of the Sacramento Valley due to the influence of the
ocean winds through the Delta ~\citep{espe16a}. These three cultivars
have a higher degree of cold tolerance, high yield potential, and
relatively short duration and are thus broadly grown with over 70\% of
the area in these cultivars at any given time during the study
period~\citep{ucce}. There are several cultivars that were found to
have better than average yields, even after correcting for yield
declines over time. However, these cultivars (M-201, M-204, M-205, and possibly
M-209; Fig. \ref{fig:ids_comp}) are not widely adopted. Importantly,
not all medium grain cultivars that were released were purported to
have higher yields than previous cultivars. Some cultivars had special
traits, but upon release had similar yields to more broadly adapted
cultivars. Examples of this include M-208, which had resistance to
blast disease (\textit{Magnaporthe grisea}) and M-402 which is a
premium quality cultivar. Other cultivars such as M-205 and M-209,
while high yielding, do not have the cold tolerance of the other
cultivars and are thus only recommended for the northern portion of
the Sacramento Valley.

@@

The high yielding variety M-202 was released in 1985 and became the
most widely planted cultivar for two decades~\citep{mckenzie2014}. At
the time of release, M-202 yields were reported at 11.0 Mg
ha\tsups{-1}~\citep{m202} which is almost identical to what we . This
value of 11.0 Mg ha\tsups{-1} is what we also found in this analysis
based on the average experimental yields prior to public release
(Fig. \ref{fig:first_yrs}). Since the release of M-202, the rice
yields of newly released varieties cultivars have remained between
10.8 and 11.5 Mg ha\tsups{-1} (Fig. \ref{fig:first_yrs}), with the
exception of M-204 and M-205 (M-204 is no longer grown and M-205 has
limited adoaptation range so is only grown in a limited area). The
variety cultivar M-206 was released in 2004 and is currently still the
most widely grown rice varietycultivar in California. At the time of
its release, M-206 was reported to yield \Sexpr{round(9350*1.12/1000,1)} Mg
ha\tsups{-1} -- similar to what we found based on our analysis
(Fig. \ref{fig:first_yrs}). Based on this premise, it is clear that
improvements in yield potential have not been made over the 33-yr
period of this study. In contrast, both~\citet{Omar2014} and
~\citet{mckenzie2014}, report yield potential gains in California rice
systems based on analyses that showed that newly released cultivars
had higher yields than previously released cultivars. Based on their
analyses,~\citet{Omar2014} reported yield gains from the rice breeding
program of 37 to 60 kg ha\tsups{-1} yr\tsups{-1} and
~\citet{mckenzie2014} a yield increase of 25 kg ha\tsups{-1}
yr\tsups{-1}. The simpliest model (Model 2) in this study is similar
to the analyses of~\citet{Omar2014} and ~\citet{mckenzie2014}, and
results in similar findings; yields were estimated to increase by
\Sexpr{only_release_coef} kg ha\tsups{-1} yr\tsups{-1}
(Fig. \ref{fig:wo_erosion}), almost identical to estimates by
~\citet{mckenzie2014}.

The only way to reconcile the two apparently contradictory findings above
is that the yield of potential of cultivars decline over
time such that when newly released cultivars are compared
with the older cultivar (as shown by ~\citet{Omar2014},
~\citet{mckenzie2014}, and Fig. \ref{fig:wo_erosion}), yields of the
new cultivar are higher. Here we provide evidence that
cultivar yield does decline - on average at a rate of
\Sexpr{yrs_coef} kg ha\tsups{-1} yr\tsups{-1}
(Fig. \ref{RES_year}). ~\citet{Peng2010} also found that a cultivar's
yield potential declines over time and reported a 15\%
decline over a 30 yr period.  In this study, over a 30 year period and
a \Sexpr{yrs_coef} kg ha\tsups{-1} yr\tsups{-1} yield decline rate,
the yield decline would be approximately
\Sexpr{round((yrs_coef*-30)/mean(mod_data$yield_kg), 3)*100}\%. Given
the estimates here using the model which accounts for these declines
suggests that there is only a
\Sexpr{round(sum((raw_samples$release_yr_c*scl)>=25)/nrow(raw_samples)*100,1)}\%
probability of the rate of yield increase is equal to or greater than
25 kg\tsups{-1} ha\tsups{-1} year\tsups{-1}.

This analysis suggests that the CCRRF breeding program has
maintained but not substantially improved the yield potential in the
face of changing biotic and abiotic pressures over the 33 year time
frame of this study. This conclusion holds
true even when evaluating the cultivars which have been breed for
broad adaption and high yield potential (M-202, M-206, and
M-105).

@@

Yields of a particular cultivar may decline over time for a number of
reasons. First, ~\citet{Nie2009} reported yield declines in rice due
to continuously flooded conditions over a long time period which could
be overcome with improved N management. This is not likely the case
here as the fields used in this study grow rice only one season a year
and tend to be non-flooded for the remainder of the year. Furthermore,
at this location and unlike most CA rice fields, the fields are left
fallow every other year. Secondly, small changes in biotic and abiotic
forces have previously been speculated as the cause for yield decline
over time of the cultivar IR8 at IRRI~\citep{Peng1998, Peng2000,
  Peng2010}. Since the cultivar's genetics remain stable from the time
of selection but the conditions the cultivar experiences over time are
continuously changing, accumulated changes too small to reliably
measure could plausibly lead to yield declines over
time~\citep{Peng1998, Peng2000, Peng2010, Mackay2011}. However, it is
not clear if this mechanism is responsible for the yield decline over
time of \Sexpr{yrs_coef} Mg ha\tsups{-1} year\tsups{-1} found in this
analysis. Biotic stressors such as pests and diseases are low in this
Mediterranean climate. Furthermore, these fields are fallowed every
other year and the residues are burned, further lowering the potential
for pest and disease pressure. Thirdly, changing climate may influence
yield potential with either warmer~\citep{van2013} or
cooler~\citep{espe2017} temperatures reducing
yields. Yield potential can be defined as the yield of optimum
cultivar in the absence of biotic and abiotic stresses in a given
environment \citep{vanIttersum2013}. However, ~\citet{espe2017} found
no substantial trends in growing season temperatures (either high,
low, or average temps) in the Sacramento Valley during the study
period, suggesting that older cultivars do not have lower yield
potential due to changing climatic factors. Our analysis
offers inconclusive evidence of an interaction between cultivar and
year (i.e., the CI for the interactions overlapped with each other and zero), visual examination suggests that cultivars may be
different in terms of rates of yield decline
(Fig. \ref{fig:all_cult}); however, we do not have sufficient data nor
the proper controls in the existing data set to fully test this
hypothesis. More data is needed to further explore the exact cause of
the yield declines observed here, as well to determine if different
cultivars have differential declines over time.
